---
title: "Introduction to phoenics"
author: "Camille Guilmineau, Rémi Servien et Nathalie Vialaneix"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Introduction to phoenics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`phoenics` is an **R** package designed to perform a differential analysis for
metabolomics datasets, while accounting for metabolomic pathway information.
More precisely, `phoenics` performs a test at pathway level based on metabolite 
quantifications and information on pathway metabolite composition. The 
automatic query of metabolic pathways is implemented in the package. See
[Guilmineau *et al.*, 2025](#guilmineau).

```{r loadLib, message=FALSE}
library("phoenics")
```

# MTBLS422 dataset

The dataset `MTBLS422` contains data obtained from the study 
[Choo *et al.*, 2017](#choo). Raw data have been made available on MetaboLights 
(with the id MTBLS422 https://www.ebi.ac.uk/metabolights/editor/MTBLS422). 
Metabolite quantifications were obtained based on the raw signal using 
[ASICS](https://www.bioconductor.org/packages/release/bioc/html/ASICS.html) 
package and are provided in the object `quantif` included in the dataset.
```{r loadData}
data("MTBLS422")
ls()
head(quantif)
```

They can be transformed into a data format suitable for `phoenics` with:
```{r quantif}
quantif_f <- from_ASICS_to_PHOENICS(quantif)
head(quantif_f)
```

In addition, pathway information was obtained using 
[KEGGREST](https://www.bioconductor.org/packages/release/bioc/html/KEGGREST.html) 
package and is included in `pathways`:
```{r pathways}
head(pathways)
```

The user can also query this information using the function `pathway_search`:
```{r searchPaths, eval=FALSE}
pathways <- pathway_search(metab = colnames(quantif), organism = "mmu")
```

# `phoenics` test

## Perform test

The test implemented in `phoenics` is a mixed model with fixed and random
effects. The current study design is provided in `design`:
```{r design}
head(design)
```

The `Age` and `Treatment` are thus used as fixed effects and the `Mouse` is 
used as a random effect:
```{r test, message=FALSE}
out_test <- test_pathway(quantif_f, design, pathways, 
                         fixed = c("Age", "Treatment"), random = "Mouse", 
                         npc = 2, model = "blmer")
out_test
```

## Access test results

Results are organized as a list where each element of the list corresponds to a
tested pathway:
```{r pathwayNames}
names(out_test)
```

The function `extract` can be used to extract results on a given pathway:
```{r extract}
res_1 <- extract(out_test, "Galactose metabolism")
```

Fixed effect $p$-values and full model are provided, respectively, in entries
`test_pathway` and `model` of this object.
```{r detailedRes}
res_1[[1]]$test_pathway
res_1[[1]]$model
```

In addition, adjusted $p$-values [(BH method)](#BH) across pathways can be 
directly obtained with the function `adjust_pval`:
```{r BH}
adjust_pval(out_test)
```


## Explore PCA results

PCA results can be explored using the entry PCA:
```{r}
res_1[[1]]$PCA
```

In particular, `phoenics` implements some plots that can be directly obtained
using
```{r plotPCA}
plot(out_test, pathway_id = "Galactose metabolism", plot = "var")
plot(out_test, pathway_id = "Galactose metabolism", plot = "ind", 
     habillage = "Age")
plot(out_test, pathway_id = "Galactose metabolism", plot = "eig")
plot(out_test, pathway_id = "Galactose metabolism", plot = "var")
plot(out_test, pathway_id = "Galactose metabolism", plot = "group")
```

# Options of `phoenics` main test function


## Switching from PCA to MFA

A MFA step, more suited to  can be used in place of the PCA. In this case, the
levels first fixed effect are used to split the effects into different tables 
which are matched by the levels of the first random effect (the typical case is
when the first fixed effect contains information on time points in a repeated
measurement study where samples are provided by the first random effect).

```{r testMFA}
out_test <- test_pathway(quantif_f, design, pathways, 
                         fixed = c("Age", "Treatment"), random = "Mouse", 
                         npc = 2, model = "blmer", analysis = "MFA")
out_test
```


## Wrapper to automatically query pathways

The option `pathways = "auto"` can also be used to automatically query 
KEGGREST instead of performing the two step analysis relying on 
`test_pathway`:

```{r auto, eval=FALSE}
out_test3 <- test_pathway(quantif, design, pathways = "auto", 
                          fixed = c("Age", "Treatment"), random = "Mouse", 
                          npc = 2, model = "blmer", organism = "mmu")
out_test3
```


# References

[1] <a name="guilmineau"></a> Guilmineau, C., Tremblay-Franco, M., 
Vialaneix, N., & Servien, R. (2025). phoenics: a novel statistical approach for
longitudinal metabolomic pathway analysis. *BMC Bioinformatics*, **26**: 105.
DOI: [10.1186/s12859-025-06118-z](http://dx.doi.org/10.1186/s12859-025-06118-z).

[2] <a name="choo"></a> Choo J. M., Kanno T., Zain N. M. M., Leong L. E. X., 
Abell G. C. J., Keeble J. E., Bruce K. D., Mason A. J., Rogers G. B. (2017). 
Divergent relationships between fecal microbiota and metabolome following 
distinct antibiotic-induced disruptions. *mSphere*, **2**(1).
DOI: [10.1128/msphere.00005-17](https://dx.doi.org/10.1128/msphere.00005-17).

[3] <a name="BH"></a> Benjamini Y, Hochberg Y (1995). Controlling the false 
discovery rate: a practical and powerful approach to multiple testing. 
*Journal of the Royal Statistical Society, Series B*, **57**(1): 289–300. 
DOI: [10.1111/j.2517-6161.1995.tb02031.x](https://dx.doi.org/10.1111/j.2517-6161.1995.tb02031.x)


# Session information

```{r session}
sessionInfo()
```